---
title: "Exploration"
output:
  pdf_document: default
  html_document: default
---

# Lyft Data Exploration

Here we explore the data I've collected while driving for Lyft. To record this data, I have a Google Spreadsheet which I keep open in the background of my phone while I'm out driving. I record the start and end time of each ride, in addition to my odometer reading and gas usage (expressed in decimal gallons). It took a dozen rides or so to calibrate my process, but now I have a flow worked out so I only record the necessary information while I'm stopped.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Libraries

```{r}
library(colorspace)
library(tidyverse)
library(gridExtra)
library(boot)
```

## Load Data and Format

I also have kept track of all the gasoline I've purchased (kept in a separate Google Spreadsheet). Here is the weighted (by gallons purchased) average price of gas. 

```{r}
gasPrice = 3.5006 # $ / gal
```

Next, we load the data. The difference between `data_all` and `data_drv` is that `data_all` includes *everything*, including all driving that takes place when I don't have any passengers (e.g. when I'm searching for a ride). For the most part, `data_drv` is of most importance.

```{r}
data_all = read_csv("CleanLyftData_All.csv", col_types = cols())

data_all$DOW = factor(data_all$DOW, levels = seq(1,7), labels = c("Mon", "Tue", "Wed", "Thur", "Fri", "Sat", "Sun"))
data_all$Month = factor(data_all$Month, levels = seq(1,12), labels = seq(1,12))
data_all$StartLocation = factor(data_all$StartLocation)
data_all$Period = factor(data_all$Period)
data_all$Movement = factor(data_all$Movement)
data_all$Origin = factor(data_all$Origin)
data_all$Goal = factor(data_all$Goal)
bins = seq(6, 22, 2)
data_all$StartTimeBin = cut(data_all$StartTime, breaks = c(bins, 24), labels = bins)

data_drv = read_csv("CleanLyftData_Drives.csv", col_types = cols())

data_drv$DOW = factor(data_drv$DOW, levels = seq(1,7), labels = c("Mon", "Tue", "Wed", "Thur", "Fri", "Sat", "Sun"))
data_drv$Month = factor(data_drv$Month, levels = seq(1,12), labels = seq(1,12))
data_drv$StartLocation = factor(data_drv$StartLocation)
data_drv$Period = factor(data_drv$Period)
data_drv$Movement = factor(data_drv$Movement)
data_drv$Origin = factor(data_drv$Origin)
data_drv$Goal = factor(data_drv$Goal)
data_drv$StartTimeBin = cut(data_drv$StartTime, breaks = c(bins, 24), labels = bins)


paste(nrow(data_drv), "drives")

data_all %>%
  filter(Session > max(data_drv$Session) - 3) %>%
  group_by(Session) %>%
  summarise(Wage = paste("$", round(sum(Earnings + Tips, na.rm = T) * (60 / sum(Duration)), 2), sep=""),
            AdjWage = paste("$", round(sum(Earnings + Tips - GasUsage * gasPrice, na.rm = T) * (60 / sum(Duration)), 2), sep=""),
            Revenue = sum(Earnings + Tips, na.rm = T),
            GasCost = round(sum(GasUsage) * gasPrice, 2)) %>%
  as.matrix()
```

## Totals

Let's view a handful of interesting summary statistics.

```{r}
c("Total Distance" = sum(data_drv$AdjDistance, na.rm = T), 
  "Total Hours" = sum(data_drv$AdjDuration, na.rm = T) / 60,
  "Drives per Session" = nrow(data_drv) / max(data_drv$Session))
```

```{r}
c("Total Days" = as.integer(max(data_drv$Date) - min(data_drv$Date)),
  "Total Passengers" = sum(data_drv$Passengers),
  "Mean Hours per Week" = round((sum(data_drv$AdjDuration, na.rm = T) / 60) / (as.integer(max(data_drv$Date) - min(data_drv$Date)) / 7), 2),
  "Mean Drives per Week" = round(nrow(data_drv) / (as.integer(max(data_drv$Date) - min(data_drv$Date)) / 7), 2))
```

```{r}
c("Mean Revenue per Drive" = round(sum(data_drv$Earnings + data_drv$Tips) / nrow(data_drv), 2),
  "Mean Gas Expenditure per Drive" = round(sum(data_drv$GasUsage * gasPrice) / nrow(data_drv), 2),
  "Mean Gas Expenditure per Session" = round(sum(data_drv$GasUsage * gasPrice) / max(data_drv$Session), 2))
```

What's the median tip amount among those who tip?

```{r}
quantile(data_drv[data_drv$Tips > 0, ]$Tips, 0.5)
```

### Preview Data

```{r}
tail(data_drv)
```

## Visualizations

Just a short cut for style.

```{r}
theme = theme_minimal()
```

### Wage Distribution

```{r}
maint = (600 + 30 + 100 + 300) / 5000 # depreciation + oil + service + parts per the next 5,000 miles
```

Next we'll create a few measures of earnings, becoming more specific in terms of what's included as the list goes down. The last measure is likely the most realistic of what I earn. 95% boostrapped Confident Intervals are listed as well.

```{r}
b = boot((data_drv$Earnings + data_drv$Tips) * (60/data_drv$Duration), function (v, ix) mean(v[ix]), R = 5000)
ci = boot.ci(b, type = "perc")$perc[1, ]
sprintf("$%0.2f (95%% CI: $%0.2f — $%0.2f)", b$t0, ci[4], ci[5])

b = boot(data_drv$Earnings * (60/data_drv$Duration), function (v, ix) mean(v[ix]), R = 5000)
ci = boot.ci(b, type = "perc")$perc[1, ]
sprintf("$%0.2f (95%% CI: $%0.2f — $%0.2f)", b$t0, ci[4], ci[5])

b = boot((data_drv$Earnings + data_drv$Tips) * (60/data_drv$AdjDuration), function (v, ix) mean(v[ix]), R = 5000)
ci = boot.ci(b, type = "perc")$perc[1, ]
sprintf("$%0.2f (95%% CI: $%0.2f — $%0.2f)", b$t0, ci[4], ci[5])

b = boot((data_drv$Earnings + data_drv$Tips - data_drv$AdjGas * gasPrice) * (60/data_drv$AdjDuration), function (v, ix) mean(v[ix]), R = 5000)
ci = boot.ci(b, type = "perc")$perc[1, ]
sprintf("$%0.2f (95%% CI: $%0.2f — $%0.2f)", b$t0, ci[4], ci[5])

b = boot((data_drv$Earnings + data_drv$Tips - data_drv$AdjGas * gasPrice - maint * data_drv$AdjDistance) * (60/data_drv$AdjDuration), function (v, ix) mean(v[ix]), R = 5000)
ci = boot.ci(b, type = "perc")$perc[1, ]
sprintf("$%0.2f (95%% CI: $%0.2f — $%0.2f)", b$t0, ci[4], ci[5])
```

### Driving Days and Times

How do earnings correlate with time and day of the ride?

```{r,}
p1 = ggplot(data_drv, aes(StartTime, AdjWage)) + theme +
  geom_point(color = "#EA0B8C") +
  geom_smooth(method = mgcv::gam, formula = y ~ s(x, bs = "gp", k = 10), se = F, color = "black") +
  labs(title = "Adjusted Wage by Start Time", x = "Start Time", y = "Adjusted Wage") +
  scale_x_continuous(breaks = seq(4, 24, 2))

res = data.frame()

for (dow in levels(data_drv$DOW)) {
  
  f = function(data, indices) {
    return(mean(data[indices]))
  }
  
  d = data_drv[data_drv$DOW == dow, ]
  b = boot(d$AdjWage, statistic = f, R = 1000)
  
  r = boot.ci(b, type = "bca")$bca
  
  res[dow, c("lb", "AdjWage", "ub")] = list(r[1, 4], b$t0, r[1, 5])
}

p2 = rownames_to_column(res, "DOW") %>%
  ggplot() + theme + 
  geom_col(aes(DOW, AdjWage, fill = AdjWage)) +
  geom_errorbar(aes(DOW, ymin = lb, ymax = ub, width=0.2)) +
  labs(title = "Adjusted Wage by Day of the Week", y = "Adjusted Wage", x = "Day of Week") + 
  scale_x_discrete(limits = levels(data_drv$DOW)) +
  scale_fill_continuous_sequential(palette = "Magenta", begin = 0.1, end = 0.9, rev = T)

p = grid.arrange(p1, p2, ncol=2)

ggsave("StartTime_DOW.png", p, width = 10, height = 6)

ggplot(data_drv) +
  geom_tile(aes(x = StartTime, y = DOW, fill = AdjWage)) +
  labs(x = "Start Time", y = "Day of Week", fill = "Adjusted Wage", title = "Adjusted Wage per Day and Hour") +
  scale_fill_continuous_sequential(palette = "Magenta", begin = 0.1, end = 0.9, rev = T)
```

Effect of day on wage?

```{r}
summary(aov(AdjWage ~ DOW, data = data_drv))
```

### Average Wage given Days and Times

Look at the big picture of day and time on wage.

```{r}
p = data_drv %>%
  group_by(StartTimeBin, DOW) %>%
  summarise(m = sum(Earnings + Tips - gasPrice * AdjGas) * (60 / sum(AdjDuration))) %>%
  ggplot() + theme +
  geom_tile(aes(x = StartTimeBin, y = DOW, fill = m)) +
  labs(x = "Start Time", y = "Day of Week", fill = "Adjusted Wage", title = "Adjusted Wage per Day and Hour") +
  scale_fill_continuous_sequential(palette = "Magenta", begin = 0.1, end = 0.9, rev = T)

p

ggsave("WageDayandTime.png", p, width = 6, height = 4)
```

### Duration on Wage

Does Lyft pay what it claims to? Do longer rides pay more or less than shorter rides?

```{r}
summary(MASS::rlm(Earnings ~ Distance + Duration, data = d))
summary(lm(Earnings ~ Distance + Duration, data = d))
```

```{r}
miles = 0.6525
hours = 0.225 * 60
```

```{r}
data_drv$Speed = data_drv$Distance / (data_drv$Duration/60)
```

```{r}
summary(lm(Speed ~ log(Distance), data = data_drv))
```

```{r}
D = seq(1, 40, 0.5)

W = (9.234 + 9.377 * log(D)) * 0.6525 + 13.5

p = ggplot() + theme +
  geom_line(aes(D, W)) +
  geom_label(aes(x = 30, y = 35, label = "W == (beta[0] + beta[1] %*% log(D)) %*% L + H"), parse = T) +
  labs(x = "Distance (miles)", y = "Wage ($ / hr)", title = "Wage as a function of distance")

p

ggsave("WageDistance.png", p, width = 5, height = 3)
```

### Time Labels

How are my artificial time labels correlated with wage? For the rule set, see the `CleanData.R`.

```{r}
table(data_drv$TimeLabel)
```

Do nightlife rides pay more than non-nightlife rides, as some online articles claim?

```{r}
t.test(data_drv[data_drv$TimeLabel == "Nightlife", ]$AdjWage, data_drv[data_drv$TimeLabel != "Nightlife", ]$AdjWage, conf.level = 0.95)
```

Are there differences between commutes?

```{r}
d = data_drv[data_drv$TimeLabel %in% c("MorningCommute", "AfternoonCommute", "EveningCommute"), ]
anova(lm(AdjWage ~ TimeLabel, data = d))
TukeyHSD(aov(lm(AdjWage ~ TimeLabel, data = d)))
```

View all the labels are their wages.

```{r}
res = data.frame()

for (tl in unique(data_drv$TimeLabel)) {
  
  f = function(data, indices) {
    return(mean(data[indices]))
  }
  
  b = boot(data_drv[data_drv$TimeLabel == tl, ]$AdjWage, statistic = f, R = 1000)
  
  r = boot.ci(b, type = "bca")$bca
  
  res[tl, c("lb", "est", "ub")] = list(r[1, 4], b$t0, r[1, 5])
}

rownames_to_column(res, "TimeLabel") %>%
  ggplot() + theme + 
  geom_bar(aes(TimeLabel, est, fill = est), stat="identity") +
  geom_errorbar(aes(TimeLabel, ymin = lb, ymax = ub, width=0.1)) +
  labs(title = "Wage given Time Label") +
  scale_fill_continuous_sequential(palette = "Magenta", begin = 0.1, end = 0.9, rev = T)
```

### Drive Position

Do the first and last drives of a session differ in wage from the rides that come in-between?

```{r}
res = data.frame()

for (pos in unique(data_drv$Position)) {
  
  f = function(data, indices) {
    return(mean(data[indices]))
  }
  
  b = boot(data_drv[data_drv$Position == pos, ]$AdjWage, statistic = f, R = 1000)
  
  r = boot.ci(b, type = "bca")$bca
  
  res[pos, c("lb", "AdjWage", "ub")] = list(r[1, 4], b$t0, r[1, 5])
}

p = rownames_to_column(res, "Position") %>%
  ggplot() + theme + 
  geom_col(aes(Position, AdjWage, fill = AdjWage)) +
  geom_errorbar(aes(Position, ymin = lb, ymax = ub, width=0.1)) +
  labs(title = "Adjusted Wage given Drive Position", y = "Adjusted Wage") +
  scale_x_discrete(limits = c("First", "Middle", "Last")) +
  scale_fill_continuous_sequential(palette = "Magenta", begin = 0.1, end = 0.9, rev = T)

p

ggsave("WagePosition.png", p, width = 6, height = 4)
```

It appears as though the last ride is likely to earn less than the other two types of rides.

```{r}
summary(aov(AdjWage ~ Position, data = data_drv))
TukeyHSD(aov(AdjWage ~ Position, data = data_drv))
```

### Drive Ratings

Only recently have I started to log satisfaction of the drive. More on this to come.

```{r}
data_drv %>%
  group_by(StartTimeBin, DOW) %>%
  summarise(m = mean(RatingConversation, na.rm = T)) %>%
  complete(DOW, fill = list(m = 0)) %>%
  ggplot() + theme +
  geom_tile(aes(x = StartTimeBin, y = DOW, fill = m)) +
  labs(title = "Mean Drive Rating given Days and Times")
```

## Start Location

What are the top 3 starting locations? How many starting locations have only been visited once? How many total are there?

```{r}
SL.counts = data_drv %>%
  group_by(StartLocation) %>%
  summarise(n = n()) %>%
  arrange(desc(n))

head(SL.counts, 3)

sum(SL.counts$n == 1)
nrow(SL.counts)
```

How does where the ride is started correlate with how much I'll earn?

```{r}
res = data.frame()

for (loc in unique(data_drv$StartLocation)) {
  
  f = function(data, indices) {
    return(mean(data[indices]))
  }
  
  d = data_drv[data_drv$StartLocation == loc, ]$AdjWage
  
  if (length(d) >= 5) {
    b = boot(data_drv[data_drv$StartLocation == loc, ]$AdjWage, statistic = f, R = 1000)
    
    r = boot.ci(b, type = "bca")$bca
    
    res[loc, c("lb", "AdjWage", "ub")] = list(r[1, 4], b$t0, r[1, 5])
  }
}

d = rownames_to_column(res, "StartLocation")

p = d %>%
  ggplot() + theme + 
  geom_col(aes(StartLocation, AdjWage, fill = AdjWage)) +
  geom_errorbar(aes(StartLocation, ymin = lb, ymax = ub, width=0.2)) +
  scale_x_discrete(limits = d[order(d$AdjWage), "StartLocation"]) +
  scale_fill_continuous_sequential(palette = "Magenta", begin = 0.1, end = 0.9, rev = T) +
  coord_flip() +
  labs(title = "Adjusted Wage given Start Location", x = "Start Location", y = "Adusted Wage")

p

ggsave("StartLocationWage.png", p, width = 10, height = 6)
```

## Wait Times

First, filter a new dataset to only wait times between rides.

```{r}
wd = data_all %>% filter(Period == "Search" & Movement != "Home")
```

```{r}
library(mgcv)
```

Day of the week does not seem to be correlated with whether I'll have to wait between rides.

```{r}
summary(aov(glm(I(Duration == 0) ~ DOW, data = wd, family = "binomial")))
```

Day of the week does not seem to be correlated with how much I'll have to wait between rides.

```{r}
summary(aov(glm(Duration ~ DOW, data = wd[wd$Duration > 0, ], family = Gamma(link = "log"))))
```

```{r}
p = data_all %>%
  filter(Period == "Search" & Movement != "Home") %>%
  group_by(StartTimeBin, DOW) %>%
  summarise(m = mean(Duration)) %>%
  ggplot() + theme +
  geom_tile(aes(x = StartTimeBin, y = DOW, fill = m)) +
  labs(x = "Start Time", y = "Day of Week", fill = "Wait Time", title = "Wait Times per Day and Hour") +
  scale_fill_continuous_sequential(palette = "Magenta", begin = 0.1, end = 0.9, rev = T)

p

ggsave("WaitDayandTime.png", p, width = 6, height = 4)
```

Wait Times for Location

```{r}
res = data.frame()

mask = data_all$Period != "Drive"

for (loc in unique(data_all[mask, ]$StartLocation)) {
  
  f = function(data, indices) {
    return(mean(data[indices]))
  }
  
  d = data_all[mask & data_all$StartLocation == loc, ]$Duration
  
  if (length(d) > 5) {
    b = boot(data_all[mask & data_all$StartLocation == loc, ]$Duration, statistic = f, R = 1000)
    
    r = boot.ci(b, type = "bca")$bca
    
    res[loc, c("lb", "Duration", "ub")] = list(r[1, 4], b$t0, r[1, 5])
  }
}

d = rownames_to_column(res, "StartLocation")

p = d %>%
  ggplot() + theme + 
  geom_col(aes(StartLocation, Duration, fill = Duration)) +
  geom_errorbar(aes(StartLocation, ymin = lb, ymax = ub, width=0.2)) +
  scale_x_discrete(limits = d[order(d$Duration), "StartLocation"]) +
  scale_fill_continuous_sequential(palette = "Magenta", begin = 0.1, end = 0.9, rev = T) +
  coord_flip() +
  labs(title = "Wait Time given Start Location", x = "Start Location", y = "Duration")

p
```

## Determinants

Does having conversation with the passenger make a difference on how much they tip? What if it's a shared ride? What about their interaction?

```{r}
mean(data_drv$Tips > 0)
mean(data_drv[data_drv$Tips > 0, ]$Tips)
```

```{r}
p = data_drv %>%
  filter(Tips > 0) %>%
  group_by(Conversation, Shared) %>%
  summarise(avg = mean(Tips)) %>%
  ggplot() + theme +
  geom_point(aes(Conversation, avg, color = Shared, shape = Shared), size=4) +
  geom_line(aes(Conversation, avg, group = Shared), lty=2) +
  lims(y = c(0, 4)) +
  labs(x = "Conversation", y = "Mean Tip", title = "Tips by Shared and Conversation", subtitle = "For Drives that left a Tip > $0")

p

ggsave("TipsInteraction.png", p, width = 6, height = 5)
```

```{r}
tip.model = glm(Tips ~ Distance + Duration + Passengers + Goal + Origin, data = data_drv %>% filter(Tips > 0), family = Gamma(link = "log"))

tip.model.c = glm(Tips ~ Conversation + Distance + Duration + Passengers + Goal + Origin, data = data_drv %>% filter(Tips > 0), family = Gamma(link = "log"))
a = anova(tip.model, tip.model.c, "Chisq")
a
pchisq(a$Deviance[2], a$Df[2], lower.tail = F)


tip.model = glm(Tips ~ Distance + Duration + Passengers + Goal + Origin, data = data_drv %>% filter(Tips > 0), family = Gamma(link = "log"))

tip.model.c = glm(Tips ~ Shared + Distance + Duration + Passengers + Goal + Origin, data = data_drv %>% filter(Tips > 0), family = Gamma(link = "log"))
a = anova(tip.model, tip.model.c, "Chisq")
a
pchisq(a$Deviance[2], a$Df[2], lower.tail = F)


tip.model = glm(Tips ~ Conversation + Shared + Distance + Duration + Passengers + Goal + Origin, data = data_drv %>% filter(Tips > 0), family = Gamma(link = "log"))

tip.model.c = glm(Tips ~ Conversation + Shared + Conversation:Shared + Distance + Duration + Passengers + Goal + Origin, data = data_drv %>% filter(Tips > 0), family = Gamma(link = "log"))
a = anova(tip.model, tip.model.c, "Chisq")
a
pchisq(a$Deviance[2], a$Df[2], lower.tail = F)
```

```{r}
tip.model = glm(I(Tips > 0) ~ Distance + Duration + Passengers + Goal + Origin, data = data_drv, family = "binomial")

tip.model.c = glm(I(Tips > 0) ~ Conversation + Distance + Duration + Passengers + Goal + Origin, data = data_drv, family = "binomial")
coef(tip.model.c)["ConversationTRUE"]
a = anova(tip.model, tip.model.c, "Chisq")
a
pchisq(a$Deviance[2], a$Df[2], lower.tail = F)


tip.model = glm(I(Tips > 0) ~ Distance + Duration + Passengers + Goal + Origin, data = data_drv, family = "binomial")

tip.model.c = glm(I(Tips > 0) ~ Shared + Distance + Duration + Passengers + Goal + Origin, data = data_drv, family = "binomial")
coef(tip.model.c)["SharedTRUE"]
a = anova(tip.model, tip.model.c, "Chisq")
a
pchisq(a$Deviance[2], a$Df[2], lower.tail = F)


tip.model = glm(I(Tips > 0) ~ Conversation + Shared + Distance + Duration + Passengers + Goal + Origin, data = data_drv, family = "binomial")

tip.model.c = glm(I(Tips > 0) ~ Conversation + Shared + Conversation:Shared + Distance + Duration + Passengers + Goal + Origin, data = data_drv, family = "binomial")
a = anova(tip.model, tip.model.c, "Chisq")
a
pchisq(a$Deviance[2], a$Df[2], lower.tail = F)
```
